# Deploying a packaged shiny application

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, eval = FALSE)
```

It makes sense to packages shiny applications to better control their dependencies. However, there are some requirements for the structure of the packaged shiny application to be able to be deployed. One option is to use the [`golem` package](https://cran.r-project.org/web/packages/golem/index.html), which will initialize the shiny application with its framework that does support to deploy the shiny application as a package. But sometimes you may not want to add an entire framework to an existing application and instead add this support manually. Since we did not find any good documentation of this online (as of Nov 2020), we investigated this ourselves and are happy to share our findings here.

### Entry point
The application needs an entry point which should be named `app.R` and be situated in the root of the package next to `DESCRIPTION` and `NAMESPACE`. It should contain only what is required for the entry point to be able to load the application, e.g.:
```{r, eval = F}
pkgload::load_all(export_all = FALSE, helpers = FALSE, attach_testthat = FALSE)
# PKG is the name of the packaged shiny application. This function wraps around shiny::shinyApp().
PKG::PKG_app()
```

### server and ui
- Both `server` and `ui` need to be functions in order to work in the packaged shiny application context. `server` should already be a function and it is enough to wrap `ui` into a function without any arguments or return statements.
- In the beginning of the `ui` function, we also need to add a call to `shiny::addResourcePath` to make static resources available.

### non-CRAN dependencies
Deploying a packaged shiny application which uses non-CRAN sources like Github can also cause issues. It is recommended to list these dependencies under [`Remotes:`](https://cran.r-project.org/web/packages/devtools/vignettes/dependencies.html) instead of e.g. `Imports:`, to make sure package versioning tools like `renv` notice the difference.
