# Version-stable R development with Docker


```{r setup-version-stable-r-development, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, eval = FALSE)
```

In the context of productive solutions, it is essential to have full control
over the code-base and environment to ensure reproducibility and stability of
the setup. In the case of R-based projects, this implies fixing and aligning the
version of R as well as package and system dependencies. In order to achieve
well-managed release pipelines, a key aspect is to guarantee full equivalence of
the development setup to (alternative) target productive stages.

This guide illustrates an approach to manage a version-stable R development
environment based on containerized solutions leveraging the
[Rocker project](https://www.rocker-project.org/), allowing the coexistence
of multiple dockerized development flavors, to match various target production
environments or projects.


## Version-stable deployments

When deploying R applications (e.g. a Shiny app) using Docker containers, it is
important to control versioning of R and packages for the sake of reproducibilty
and stability of the deployments. For this reason,
[version-stable](https://github.com/rocker-org/rocker-versioned) images are
provided as part of the [Rocker project](https://www.rocker-project.org/) and
used as a basis for deploying productive applications.

Each version-stable Rocker image has an associated _tag_ for all non-latest R
versions (e.g. `rocker/r-ver:3.6.1`). Besides being specific to the
corresponding version of R, each tag fixes the version of contributed packages
(by using as package repository the MRAN snapshot of the last day CRAN
distributed that R version as latest release). See
[VERSIONS.md](https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md)

The `Dockerfile` of a deployed application then defines a given version-stable
image tag to start `FROM`, e.g.
```dockerfile
FROM rocker/r-ver:3.6.1
```
See
[SmaRP/Dockerfile](https://github.com/miraisolutions/SmaRP/blob/master/Dockerfile)
for an example.


## Align local development and deployment environments

When developing and testing an app locally, it is important to ensure the
environment is aligned with the target deployment environment. This might imply
using e.g. multiple R and package versions for the local development of
different applications, which clash with the typical setup (especially on
Linux systems), where only one R version (the latest release) exists.

The idea is then to rely on the same version-stable rocker containers used for
the deployments, using a containerized versioned RStudio instance for the local
development. This is available through Rocker's [versioned
stack](https://www.rocker-project.org/images/#the-versioned-stack), so we could
use e.g. `rocker/rstudio:3.6.1`.

Note that the same version-stable instance of RStudio can be used across all
different projects for which such version is relevant. For this reason, a
sensible choice is to rely on `rocker/verse`, which adds tidyverse and
devtools to the stack, as well as properly setting up R Markdown system
dependencies TinyTeX and `pandoc`, sparing the effort of the tedious extra
install. See the specific section below about 'TinyTeX considerations'.


### Running versioned RStudio instances

Assume we want to run a containerized versioned instance of RStudio for R 3.6.1,
possibly alongside instances for other versions of R.

First of all, we need to get the image from docker-hub
```{bash pull}
docker pull rocker/verse:3.6.1
```

We then want to have a running instance on `localhost` (`127.0.0.1`), with the
following setup:

- No authentication required (local setup).
- Enable root by setting the environment variable `ROOT` to `TRUE`, so that e.g.
`sudo apt-get` can be used in RStudio.
- Use a version-specific port, e.g. `3500` for R 3.5.0, `3610` for R 3.6.1 and
so on, so that we can use `localhost` for concurrent R version instances.
- The development code of all relevant projects should live outside the
container and be shared with it (and possibly many of them), e.g. under
`~/RStudioProjects` on the host machine and `/home/rstudio/RStudioProjects` in
the container.
    - For this to work w/o [permission
    issues](https://github.com/rocker-org/rocker/wiki/Sharing-files-with-host-machine#avoiding-permission-changes-when-sharing-volumes),
    the container user (`rstudio`) must match the UID of the host user (`$UID`).
- In order for the RStudio setting to persist if the container is recreated
(e.g. after pulling a new `rocker` image), we also use a shared volume (like
`~/.rstudio-docker/3.6.1`) for the `home/rstudio/.rstudio` directory, which is
version-specific in case of multiple R versions
- If we want to use Meld via the [compareWith](https://github.com/miraisolutions/compareWith/) addins, we need to
    - map the `DISPLAY` environment variable and volume `/tmp/.X11-unix`
    - add `DISPLAY` to `Renviron`
    - install Meld
    - install `dbus-x11`
- Use a version-specific name for the container running the RStudio instance,
e.g. `rstudio_3.6.1`.


```{bash run}
R_VER=3.6.1
SHARED_DIR=RStudioProjects
docker run -d --restart=always \
  -p 127.0.0.1:$(echo $R_VER | sed 's/[.]//g')0:8787 \
  -e DISABLE_AUTH=true \
  -e ROOT=TRUE \
  -e USERID=$UID \
  -v $HOME/$SHARED_DIR:/home/rstudio/$SHARED_DIR \
  -v $HOME/.rstudio-docker/$R_VER:/home/rstudio/.rstudio \
  -e DISPLAY=$DISPLAY \
  -v /tmp/.X11-unix:/tmp/.X11-unix:ro \
  --name rstudio_$R_VER \
  rocker/verse:$R_VER
# R and RStudio are not getting the DISPLAY environment variable
docker exec rstudio_$R_VER bash -c \
  'echo "DISPLAY=${DISPLAY}" >> /usr/local/lib/R/etc/Renviron'
# Install Meld
docker exec rstudio_$R_VER bash -c \
  'apt-get update && apt-get install -y --no-install-recommends meld dbus-x11'
```
The running RStudio can then be accessed by visiting `http://localhost:3610/`.

You may find convenient to define a shell function as follows

```{bash run_rstudio_ver-def}
run_rstudio_ver() {
  local R_VER=${1:?"you must supply the R version as first argument"}
  local SHARED_DIR=${2:?"you must supply the shared directory as second argument"}
  local RVER_IMAGE=${3:-"verse"}
  local BASE_IMAGE=rocker/$RVER_IMAGE:$R_VER
  local PORT=$(echo $R_VER | sed 's/[.]//g')0
  local CONTANER_NAME=rstudio_$R_VER
  echo "Containerized version-stable RStudio for R "$R_VER\
       "based on image "$BASE_IMAGE\
       "with shared volume "$SHARED_DIR
  docker pull $BASE_IMAGE &&
  docker run -d --restart=always \
    -p 127.0.0.1:$PORT:8787 \
    -e DISABLE_AUTH=true \
    -e ROOT=TRUE \
    -e USERID=$UID \
    -v $HOME/$SHARED_DIR:/home/rstudio/$SHARED_DIR \
    -v $HOME/.rstudio-docker/$R_VER:/home/rstudio/.rstudio \
    -e DISPLAY=$DISPLAY \
    -v /tmp/.X11-unix:/tmp/.X11-unix:ro \
    --name $CONTANER_NAME \
    $BASE_IMAGE &&
  # R and RStudio are not getting the DISPLAY environment variable
  docker exec $CONTANER_NAME bash -c \
    'echo "DISPLAY=${DISPLAY}" >> /usr/local/lib/R/etc/Renviron' &&
  # Install Meld
  docker exec $CONTANER_NAME bash -c \
    'apt-get update && apt-get install -y --no-install-recommends meld dbus-x11' &&
  echo "RStudio running in container "$CONTANER_NAME" on port "$PORT &&
  echo "visit http://localhost:"$PORT
}
```

which you can re-use as compact command for any R version as follows
```{bash run_rstudio_ver-use}
run_rstudio_ver 3.6.1 RStudioProjects
```

Note that `--restart=always` specifies that the container should stay up and restart
itself after stopping, e.g. upon machine reboot or docker upgrade, so that it is
always available. Still, you can explicitly stop the running container with
```{bash stop}
docker stop rstudio_3.6.1
```
Alternatively, you can omit `--restart=always` and explicitly start the
container whenever needed with
```{bash start}
docker start rstudio_3.6.1
```

Note that `start`/`stop` operations do not affect the persistence of files
created outside the shared location, including global RStudio options such as
dark theme, diagnostic, etc. (set via _Tools > Global Options..._). On the other
hand, these files and settings do ~~not~~ (see above) persist removing the
container (`docker rm`, see below) .


### TinyTeX considerations

#### `pdfcrop`

Older `rocker/verse` images might not include `pdfcrop`, which is required for
the default and desirable cropping of PDF figures with R Markdown (see
[rocker-org/rocker-versioned#146](https://github.com/rocker-org/rocker-versioned/issues/146)).
Make sure `pdfcrop` is installed by running at R console
```{r pdfcrop}
tinytex::tlmgr_install("pdfcrop")
```
(see [R Markdown: The Definitive
Guide](https://bookdown.org/yihui/rmarkdown/pdf-document.html#figure-options-1))

#### Align TinyTeX to current remote repo

**NOTE** - This should never be needed with recent `rocker/verse` images, where
a version-stable Tex Live repository is used for the TinyTeX install (see
[rocker-org/rocker-versioned#169](https://github.com/rocker-org/rocker-versioned/issues/169)).

If you are using LaTeX and start seeing errors like
```
Remote repository is newer than local (2018 < 2019)
```
it means that you have to re-install TinyTeX. This happens e.g. with
`rocker/verse:3.6.1`, since it was build at the end of 2018 but the current
Tex Live repo is 2019. You can fix this via a **user-specific** re-installation of
TinyTeX for R. **NOTE** however that this will uninstall the system-level
TinyTeX pre-installed in `rocker/verse`.

First, make sure `/home/rstudio/bin` is part of the `PATH` environment variable.
Check this by running
```{bash check-path}
docker exec --user rstudio rstudio_3.6.1 R --slave -e 'Sys.getenv("PATH")'
```
If you don't see `/home/rstudio/bin`, you can make sure it is part of the `PATH` for R via
```{bash set-path}
docker exec --user rstudio rstudio_3.6.1 sh -c 'echo "PATH=$HOME/bin:\${PATH}" >> $HOME/.Renviron'
# check again
docker exec --user rstudio rstudio_3.6.1 R --slave -e 'Sys.getenv("PATH")'
```

Then, from the running RStudio, run
```{r reinstall-tinytex}
tinytex::reinstall_tinytex()
```


### Cleanup

```{bash cleanup}
docker rm $(docker stop rstudio_3.6.1)
```


## References

- [The Rocker Project](https://www.rocker-project.org/)
- [Shared Volumes](https://www.rocker-project.org/use/shared_volumes/)
- [Managing Users](https://www.rocker-project.org/use/managing_users/)
- [Managing containers](https://www.rocker-project.org/use/managing_containers/)
- [Rocker Wiki](https://github.com/rocker-org/rocker/wiki)
- [Sharing files with host machine](https://github.com/rocker-org/rocker/wiki/Sharing-files-with-host-machine)
