---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# roxygenExPkg

An example package showing motivated guidelines about structuring the
**roxygen2** documentation content.


## Installation

#### Local checkout

In order to install the package from a git checkout, install as follows:

```{r checkout-install, eval = FALSE}
devtools::install()
```

#### GitHub

In order to access *private repositories* on GitHub programmatically, the best
way is to generate a personal access token (PAT) online
https://github.com/settings/tokens. Once generated, you need to copy the value
and save it somewhere. Once you move away from the page, you will not be able to
retrieve the value online anymore. To make the PAT available to R and RStudio
sessions, the preferred option is to set the `GITHUB_PAT` environment variable
in your `.Renviron` (e.g. under /etc/R on a typical Linux setup) to the value
stored above. The package `devtools` will look for this environment variable.

```{r github-install, eval = FALSE}
devtools::install_github(
  "miraisolutions/techguides/roxygen_guidelines/roxygenExPkg"
)
```


## Guidelines and examples

The package contains examples, illustrating and motivating best-practices, which
can be found in the R sources of the package itself. 
The corresponding help documentation pages can be browsed to see how help content is rendered.
```{r help, eval = FALSE}
help(package = "roxygenExPkg")
```
